---
title: "Do_seq_pre_pro"
output: html_document
---

# Loading
First load required libraries and data
```{r, echo=FALSE}
library(tidyverse)
library(patchwork)
library(ggplot2)
library(grid)
library(cowplot)
library(purrr)
library(fuzzyjoin)
library(openxlsx)
library(vegan)
library(ape)
library(ggrepel)


do_0203 <- read_tsv("dojran_shotgun_data.tsv")

do_df_raw <- do_0203 %>%
  select(library, core, depth, db, everything())

#--------------- load external data of OSL dates and depths
osl_depth_ex <- read_csv("osl_raw_sorted.csv", col_names = TRUE)

do_age_depth <- read.xlsx("age_depth_sifra.xlsx", sheet = 1, colNames = TRUE) %>%
  group_by(core) %>%
  mutate(across(c(age_mean, age_median, age_2.5, age_97.5), ~ round(./1000, 2)))

# need to do some wrangling to get depth ranges ready to merge with our df's
do_osl_ages <- osl_depth_ex %>%
  group_by(core) %>%
  separate(depth_range, into = c("depth_min", "depth_max"), sep = "-", convert = TRUE) %>%
  arrange(core, depth_min)


# ----------------- join osl to do_df_raw -------------------
do_df_raw <- do_df_raw %>%
  left_join(do_age_depth, by = c('core', "depth"))%>%
  select(library, core, depth, age_mean, db, everything()) %>%
  group_by(core, depth)

unique(do_df_raw %>% filter(core == "Core DOR 4") %>% pull(age_mean))
```
# Simplifying df's
To make it easier to work we simplify the df we get from Kevin.
I'll make multiple, df_genus, df_genus_family etc. but for now only focus on genus data, so we only extract data from the df_raw which is relevant for the df_gs (genus dataframe). 
```{r, echo = FALSE}
# Define relevant columns for a genus df
columns_df_gs <- c(
  "library",
  "core",
  "depth",
  "age_mean",
  "db", 
  "phylum_name",
  "order_name",
  "family_name",
  "genus_name",
  "taxa_name", # taxa in this case refers to the reference genome which the read is mapped to, res is genus still
  "genus_count_assembly",
  "genus_deam_count",
  "sum_genus_count",
  "median_read_length_genus",
  "ancientness_genus",
  "deam5p_frac_genus",
  "deam5p_genus",
  "ref_C_sum_5p_genus",
  "deam3p_frac_genus",
  "deam3p_genus",
  "ref_C_sum_3p_genus",
  "family_count_assembly",
  "family_deam_count",
  "sum_family_count",
  "median_read_length_family",
  "ancientness_family",
  "deam5p_frac_family",
  "deam5p_family",
  "ref_C_sum_5p_family",
  "deam3p_frac_family",
  "deam3p_family",
  "ref_C_sum_3p_family",
  "age_median",
  "age_2.5",
  "age_97.5" 
  )

columns_df_gs_fm <- c(
  "library",
  "core",
  "depth",
  "age_mean",
  "db",
  "phylum_name",
  "order_name",
  "family_name",
  "genus_name",
  "taxa_name", # taxa in this case refers to the reference genome which the read is mapped to, res is genus still
  "genus_count_assembly",
  "sum_genus_count",
  "genus_deam_count",
  "median_read_length_genus",
  "ancientness_genus",
  "deam5p_frac_genus",
  "deam5p_genus",
  "ref_C_sum_5p_genus",
  "deam3p_frac_genus",
  "deam3p_genus",
  "ref_C_sum_3p_genus",
  "family_count_assembly",
  "family_deam_count",
  "sum_family_count",
  "median_read_length_family",
  "ancientness_family",
  "deam5p_frac_family",
  "deam5p_family",
  "ref_C_sum_5p_family",
  "deam3p_frac_family",
  "deam3p_family",
  "ref_C_sum_3p_family",
  "sum_phylum_count",
  "sum_order_count",
  "age_median",
  "age_2.5",
  "age_97.5" 
  )
```
# create smaller df's

```{r, echo=FALSE}
do_df_fm_gs <- do_df_raw %>% # so we define the genus data frame (df_gs) from the raw df. 
  #filter(!library %in% c("")) # This line is for removing f.e. libraries of blanks
  select(all_of(columns_df_gs_fm)) %>% # use select to only select columns relevant for df_gs 
  group_by(library, genus_name) %>%
  mutate(
    deam5p_frac_genus_family = (deam5p_genus + deam5p_family) / (ref_C_sum_5p_genus + ref_C_sum_5p_family),
    deam3p_frac_genus_family = (deam3p_genus + deam3p_family) / (ref_C_sum_3p_genus + ref_C_sum_3p_family)) %>% #add deamination rate
  filter(!sum_genus_count < 100) %>%# filter out entries with lower than 100 reads, this low reads = generally unreliable 
  slice_max(genus_count_assembly, n = 1, with_ties = FALSE) %>%
  arrange(core, depth, library)
```


#--------------------------------------- Contamination check and filter -----------------------------
## powo auto check
Here I want to make a section where we can automatically check for f.e. all the genera in our df if they are native or exotics. And check their origings. This way we can see which genera are f.e. from the Americas which would mark them as obvious modern contamination in older samples. 


```{r, echo=FALSE}
# 1 load in wcvp data (go to powo.science.kew.org to find sources)

wcvp_names <- read.table("wcvp/wcvp_names.csv", sep="|", header=TRUE, quote = "", fill=TRUE, encoding = "UTF-8")
wcvp_dist <- read.table("wcvp/wcvp_distribution.csv", sep="|", header=TRUE, quote = "", fill=TRUE, encoding = "UTF-8")

# 2 define country list of interest (TDWG level-3 area codes)
do_target_area <- c(
  "GRC", # GREECE
  "BUL", # Bulgaria
  "ALB", # Albania
  #"MCD"  # North-Macedonia does not hav TDWG level 3 code, only level 4 :(
  "TUR", # Turkey
  "YUG", # Balkan countries,
  "ITA" # Italy,
  
)

# 3 look up which genera are native from WCVP
# exclude anything non-native form df, can be used to give indication of native
do_wcvp_native <- wcvp_dist %>%
  filter(area_code_l3 %in% do_target_area,
         introduced == 0) %>%
  right_join(
    wcvp_names %>% select(plant_name_id, genus, lifeform_description),
    by = "plant_name_id"
  ) %>%
  select(genus, everything()) %>%
  filter(!is.na(genus),
         !is.na(area_code_l3)) %>%
  distinct(genus, region, area, area_code_l3, lifeform_description)

# this just gives a more detailed overview which genus is native to which country
do_wcvp_native_ov <- do_wcvp_native %>%
  mutate(native = TRUE) %>%
  pivot_wider(
    names_from = area_code_l3,
    values_from = native,
    values_fill = FALSE,
    names_prefix = "native_"
  ) %>%
  mutate(native_any = TRUE)

```

## Add native status to df
Now we use the '_wcvp_native' database we created before to add 'native' status to our seq df. 
only works for plants
```{r, echo = FALSE}
do_df_plant <- do_df_fm_gs %>%
  filter(db == "plant") %>%
  mutate(
    native = case_when(genus_name %in% do_wcvp_native$genus ~ "1",
                       TRUE ~ "0")
  ) %>%
  arrange(depth, core, library)

do_df_non_nat <- do_df_plant %>% # non native df
  filter(native == "0") %>%
  group_by(core) %>%
  arrange(core, depth)

non_native <- unique(do_df_non_nat$genus_name)

# We're pretty sure highest read scoring assembly is right one. This because all three contaminants 
do_df_corn <- do_df_plant %>%
  filter(genus_name %in% c("Zea", "Helianthus", "Nicotiana") ) %>%
  group_by(core, depth) %>%
  arrange(core, depth)

do_df_corn_ref <- do_df_corn %>%
  ungroup() %>%
  group_by(core) %>%
  summarise(
    max_reads          = max(genus_count_assembly, na.rm = TRUE),
    deam_at_max_count  = deam5p_frac_genus[which.max(genus_count_assembly)],
    genus_max_count    = genus_name[which.max(genus_count_assembly)],
    depth_max_count    = depth[which.max(genus_count_assembly)],
    
    max_deam           = max(deam5p_frac_genus, na.rm = TRUE),
    count_at_maxdeam   = genus_count_assembly[which.max(deam5p_frac_genus)],
    genus_at_maxdeam   = genus_name[which.max(deam5p_frac_genus)],
    depth_at_maxdeam   = depth[which.max(deam5p_frac_genus)]
    
    )
    
    
do_contam_colour <- c(
  "Zea"        = "#F0E143",
  "Helianthus" = "#F08E43",
  "Nicotiana"  = "#9E5F2F"
)

```

## Plot option
We want to add highest N deam rates of the LAKE cores as reference to see how our contaminations are scoring. then plot per core over depth with contaminations coloured infront. 
```{r, echo = FALSE}
# get highest read N form lakes
do_lake_highN <- do_df_fm_gs %>%
  filter(
    str_detect(core, "DOJ"),
    db == "plant",
    genus_count_assembly > 1000
  )

# get >500 reads from terrestrial
do_ter_highN <- do_df_fm_gs %>%
  filter(
    str_detect(core, "DOR"),
    db == "plant",
    genus_count_assembly > 500
  )

# Add y-lines showing approx. 1500 CE
do_ter_ages <- do_df_plant %>%
  filter(str_detect(core, "DOR")) %>%
  group_by(core, depth) %>%
  summarise(age_mean = mean(age_mean, na.rm = TRUE))

# based on that make age and depth where we have <1500 CE
do_1500_ter <- do_ter_ages %>%
  filter(age_mean > 0.500) %>%
  group_by(core) %>%
  summarise(depth_1500 = min(depth),
            age_mean = first(age_mean))
```
plot
```{r, echo = FALSE}
do_cores_corn <- unique(do_df_corn$core)

do_p_contam_list <- imap(do_cores_corn, function(this_core, core_idx) {
  
  is_leftmost  <- core_idx == 1
  is_rightmost <- core_idx == length(do_cores_corn)
  
  core_data <- do_df_corn %>% filter(core == this_core)
  
  # Get the depth and age for horizontal line
  line_info <- do_1500_ter %>%
    filter(core == this_core)
  
  line_depth <- line_info %>% pull('depth_1500')
  line_age <- line_info %>% pull('age_mean')
  
  p <- ggplot(core_data, aes(x = deam5p_frac_genus_family, y = depth)) +
    # first background points
    geom_point(data = do_ter_highN, 
               aes(x = deam5p_frac_genus_family, y = depth, fill = "ter data > 500 reads"),
               alpha = 0.5, size = 3, shape = 21, colour = "grey30") +
    geom_point(aes(fill = genus_name),
               alpha = 1, size = 3, shape = 21, colour = "black") +
    geom_hline(yintercept = line_depth, colour = "black", linetype = "dashed") +
    # Add text label with age
    annotate("text", 
             x = 0.28,  # position on x-axis (right side)
             y = line_depth, 
             label = paste0(line_age, " ka"),
             hjust = 1,  # right-align
             vjust = -0.5,  # slightly above the line
             size = 3,
             colour = "black") +
    scale_x_continuous(limits = c(0, 0.3)) +
    scale_y_reverse() +
    scale_fill_manual(values = c(do_contam_colour, "ter data > 500 reads" = "grey60")) +
    labs(
      title = this_core,
      x = "deam5p_frac_genus",
      y = if (is_leftmost) "depth" else NULL
    ) +
    theme_classic() +
    theme(
      legend.position  = "none",
      axis.text.y      = if (is_leftmost) element_text() else element_blank(),
      axis.ticks.y     = if (is_leftmost) element_line() else element_blank(),
      plot.title       = element_text(hjust = 0.5)
    )
  
  p
})

# Extract legend from 2nd core
legend_grob <- cowplot::get_legend(
  do_p_contam_list[[2]] + theme(legend.position = "right")
)

# Combine
do_p_contam_comb <- cowplot::plot_grid(
  patchwork::wrap_plots(do_p_contam_list, nrow = 1),
  legend_grob,
  ncol = 2,
  rel_widths = c(1, 0.2)
)

do_p_contam_comb


ggsave("output/do_deamr_contam_v_terr2.pdf", plot = do_p_contam_comb, width = 12, height = 10, dpi = 300)
```


##-------- Set threshold based on age -----
We added OSL ages, although fickle, we know that if any of our likely contaminants (corn, tobacco, sunflower), show up in layers older than 500 years ago, we have leaching, since these were not cultivated in Europe before that time. We use this to chek the deamination rate of these known contaminants to see if we can us them as thershold for what is a contaminant. 
```{r, echo = FALSE}
do_df_corn_leach <- do_df_corn %>%
  group_by(core) %>%
  filter(age_mean > 0.5) %>%
  arrange(core, depth)

do_contam_max_deamr <- do_df_corn_leach %>%
  group_by(core) %>%
  summarise(max_deam5p_frac_genus = max(deam5p_frac_genus, na.rm = TRUE))


##### This is the set we'll use for PCoA. we 'trust' this set ########
do_df_plant_ter <- do_df_plant %>%
  filter(core %in% c("Core DOR 2","Core DOR 3", "Core DOR 4")) 

do_df_pl_tr_cl <- do_df_plant_ter %>%  
  left_join(do_contam_max_deamr, by = "core")%>%
  group_by(core) %>%
  filter( age_mean <=0.5 | deam5p_frac_genus >= max_deam5p_frac_genus) %>%
  ungroup() %>%
  group_by(library) %>%
  arrange(core, depth)

```

#--------------------------- attempt to create deam plots like Kevin --------------------------
## Define reoccuring plot elements, colours etc
```{r, echo = FALSE}
# my_theme_min #not made yet but in future good way to work 

# Colours
kingdom_colours <- c( 
  "animals"  = "#FF401F",
  "plants"   = "forestgreen",
  "microbes" = "#FF66BE",
  "protozoa" = "#7BE0CA",
  "other"    = "#E0C27B")

db_colours <- c(
  "vertebrate_mammalian" = "#00BFC4",
  "plant"                = "#7CAE00",
  "vertebrate_other"     = "#C77CFF",
  "invertebrate"         = "#F53527",
  "protozoa"             = "cyan4",
  "bacteria"             = "#FF66BE"
  
)

```

## Try recreate Kevin deamination rate plots (standardise this later)
quick try from DS
```{r, echo = FALSE}
# Get unique cores
do_cores <- unique(do_df_raw$core)

# Separate DOR and DOJ cores (we want seperate plots for lakes and terrestrial)
dor_cores <- do_cores[grepl("DOR", do_cores)]
doj_cores <- do_cores[grepl("DOJ", do_cores)]

# Define count ranges
count_ranges <- list(
  c(0, 500),
  c(500, 1000),
  c(1000, Inf)
)

count_labels <- c("<500", "500-1000", ">1000")

# we'll create two different loops, just for simplicity sake
#first DOR (terrestrial) cores. 

do_plots <- imap(count_ranges, function(count_range, range_idx) {
  count_min <- count_range[1]
  count_max <- count_range[2]
  
  imap(dor_cores, function(this_core, core_idx) {
    
    is_leftmost <- core_idx == 1
    is_rightmost <- core_idx == length(dor_cores)
    
    #select data for specific core
    core_data <- do_df_fm_gs %>%
      filter(core == this_core,
             !db == "bacteria",
             genus_count_assembly >= count_min & genus_count_assembly < count_max)
    # select depth for hline per core
    hline_val <- do_1500_ter %>% filter(core == this_core) %>% pull()
    
    p <- ggplot(core_data, aes(x = depth, y = deam5p_frac_genus_family, fill = db, colour = db)) +
      geom_point(alpha = 1, size = 3, shape = 21, colour = "black") +
      geom_smooth(colour = "black", alpha = 0.4, se = TRUE, level = 0.95, method = "loess", span = 0.9) +
      geom_hline(yintercept = hline_val, linetype = "dashed", colour = "red") +
      scale_x_continuous(limits = c(20, 180)) +
      scale_y_continuous(limits = c(0, 0.3)) +
      scale_colour_manual(values = db_colours, limits = names(db_colours)) +
      scale_fill_manual(values = db_colours, limits = names(db_colours)) +
      labs(
        # title and axis titles based on position of plot. 
        title = if(range_idx == 1) this_core else NULL,
        x = "depth",
        y = if (is_leftmost) "deam5p_frac_genus_family" else NULL,
        # right y-axis count label
        tag = NULL
        ) +
      theme_classic() +
      theme(
        legend.position = "none",
        # y-axis left most only
        axis.text.y = if (is_leftmost) element_text() else element_blank(),
        axis.ticks.y = if (is_leftmost) element_line() else element_blank(),
        axis.title.y = if (is_leftmost) element_text() else element_blank(),
        # y-axis count right side via sec.axis trick - see below
        plot.title = element_text(hjust = 0.5)
      )
      
    # add count range label as secondary y-axis title on rightmost plot only
    if(is_rightmost) {
      p <- p + scale_y_continuous(
        limits = c(0, 0.3),
        sec.axis = sec_axis(~ ., name = count_labels[range_idx])
      ) +
        theme(
          axis.text.y.right = element_blank(),
          axis.ticks.y.right = element_blank(), 
          axis.title.y.right = element_text(angle = 90, vjust = 0.5)
        )
    }
    p
  })
  
}) %>% flatten()
 
get_legend_from <- function(plot_list, idx) {
  cowplot::get_legend(
    plot_list[[idx]] + theme(legend.position = "right")
  )
}

legend_grob <- get_legend_from(dor_plots, 3)  # grab from whatever you want

# combine all plots
do_grid <- wrap_plots(do_plots, ncol = length(dor_cores), byrow = TRUE)

dor_plots_final <- plot_grid(
  do_grid,
  legend_grob,
  ncol = 2,
  rel_widths = c(1, 0.15)
)

dor_plots_final

```

## dumb attempt
so no luck on the loop. It keeps messing up, I don't know what it is, so I'll just manually make them quickly
```{r, echo = FALSE}  
# first make df for each core seperate
do_df_dor2_fm_gs <- do_df_fm_gs %>%
  filter(core == "Core DOR 2")

do_df_dor3_fm_gs <- do_df_fm_gs %>%
  filter(core == "Core DOR 3")

do_df_dor4_fm_gs <- do_df_fm_gs %>%
  filter(core == "Core DOR 4")

do_df_dojc_fm_gs <- do_df_fm_gs %>%
  filter(core == "Core DOJ 1/24 center")

do_df_dojA_fm_gs <- do_df_fm_gs %>%
  filter(core == "Core DOJ DEL 3/24/A")

do_df_dojB_fm_gs <- do_df_fm_gs %>%
  filter(core == "Core DOJ DEL 4/24/B")

#-------------------------- Plot per core --------------------------------------------
do_dor2_deamr <- ggplot(do_df_dor2_fm_gs, aes(x = depth, y = deam5p_frac_genus, fill = db, colour = db)) +
      geom_point(alpha = 1, size = 3, shape = 21, colour = "black") +
      geom_smooth(colour = "black", alpha = 0.4, se = TRUE, level = 0.95, method = "loess", span = 0.9) +
      geom_hline(yintercept = 40, linetype = "dashed", colour = "red") +
      scale_x_continuous() +
      scale_y_continuous(limits = c(0, 0.3)) +
      scale_colour_manual(values = db_colours, limits = names(db_colours)) +
      scale_fill_manual(values = db_colours, limits = names(db_colours)) +
      labs(
        title = paste0(do_df_dor2_fm_gs$core),
        x = "depth",
        y = "deam5p_frac_genus"
        ) +
      theme_classic() 

do_dor2_deamr

```

#----------------------------- deam rate vs avg. read length------------------------------
quick plot deam_rate vs avg. read length to see if we can see any patterns.
```{r, echo = FALSE}  
do_site_colour <- c(
  "Core DOR 2" = "#EBE209",
  "Core DOR 3" = "#D6A85A",
  "Core DOR 4" = "#452A15",
  "Core DOJ 1/24 center" = "#71E362",
  "Core DOJ DEL 3/24/A" = "#30B21F",
  "Core DOJ DEL 4/24/B" = "#1B6412" 
)


# long to combine fm and gs values
do_df_long <- do_df_fm_gs %>%
  pivot_longer(
    cols = c(deam5p_frac_genus, deam5p_frac_family),
    names_to = "deam_level",
    values_to = "deam_value"
  ) %>%
  pivot_longer(
    cols = c(median_read_length_genus, median_read_length_family),
    names_to = "readlength_level",
    values_to = "readlength_value"
  ) %>%
  # keep only matching pairs (genus with genus, family with family)
  filter(
    (deam_level == "deam5p_frac_genus"  & readlength_level == "median_read_length_genus") |
    (deam_level == "deam5p_frac_family" & readlength_level == "median_read_length_family")
  ) %>%
  mutate(level = if_else(str_detect(deam_level, "genus"), "genus", "family"))

# set filters
do_pres_plot_df_di <- do_df_raw %>%
  filter(sum_genus_count > 1)

do_n_labels_di <- do_pres_plot_df_di %>%
  group_by(core) %>%
  summarise(n = n()) %>%
  mutate(label = paste0("N = ", n))

do_pres_plot_df_cl <- do_df_raw %>%
  filter(sum_genus_count > 100)

do_n_labels_cl <- do_pres_plot_df_cl %>%
  group_by(core) %>%
  summarise(n = n()) %>%
  mutate(label = paste0("N = ", n))

do_pres_plot_gs_cl <- ggplot(do_pres_plot_df_cl, aes(x = deam5p_frac_genus, y = median_read_length_genus, fill = core, colour = core)) +
  geom_point(alpha = 1, size = 3, shape = 21, colour = "black") +
  geom_smooth(colour = "red", alpha = 0.4, se = TRUE, level = 0.95, method = "lm", span = 0.9) +
  geom_text(data = do_n_labels_cl, aes(label = label), 
            x = 0.35, y = 75,  # adjust position as needed
            inherit.aes = FALSE) +
  scale_y_continuous(limits = c(30,80)) +
  scale_x_continuous(limits = c(0, 0.4)) +
  scale_colour_manual(values = do_site_colour) +
  scale_fill_manual(values = do_site_colour) +
  labs(
    title = "deam5p_frac_genus_genus vs median_read_length_genus",
    x = "deam5-_frac_genus",
    y = "median_read_length_genus"
    ) +
  facet_wrap(~core, ncol = 3) +
  theme_classic() +
  theme(legend.position = "none")

do_pres_plot_gs_cl


ggsave("output/Do_deamr_readlength_gs_clean.pdf", plot = do_pres_plot_gs_cl, width = 12, height = 10, dpi = 300)


#testest


DOR4_test <- do_df_raw %>%
  filter(core == "Core DOR 4")%>%
  summarise(depth_max = max(depth, na.rm = TRUE))

DOR4_test
  
```

## ----------- Same figure but universal taxa ---------------------
```{r, echo = FALSE}
# find taxa present in all cores
do_n_cores <- n_distinct(do_df_raw$core)

do_universal_taxa <- do_df_raw %>%
  filter(sum_genus_count > 1) %>%
  group_by(taxa_name) %>%
  summarise(cores_present = n_distinct(core)) %>%
  filter(cores_present == do_n_cores) %>%
  pull(taxa_name)

# apply filter 
do_pres_plot_uni <- do_df_raw %>%
  filter(taxa_name %in% do_universal_taxa,
         sum_genus_count > 99)

do_pres_plot_gs_uni <- ggplot(do_pres_plot_uni, aes(x = deam5p_frac_genus, y = median_read_length_genus, fill = core, colour = core)) +
  geom_point(alpha = 1, size = 3, shape = 21, colour = "black") +
  geom_smooth(colour = "red", alpha = 0.4, se = TRUE, level = 0.95, method = "lm", span = 0.9) +
  scale_y_continuous(limits = c(30,80)) +
  scale_x_continuous(limits = c(0, 0.4)) +
  scale_colour_manual(values = do_site_colour) +
  scale_fill_manual(values = do_site_colour) +
  labs(
    title = "Universal genera: deamrate vs read length",
    x = "deam5p_frac_genus",
    y = "median_read_length_genus"
  ) +
  facet_wrap(~core, ncol = 3) +
  theme_classic() +
  theme(legend.position = "none")

do_pres_plot_gs_uni

ggsave("output/Do_deamr_readlength_gs100_universal.pdf", plot = do_pres_plot_gs_uni, width = 12, height = 10, dpi = 300)
```












 
 

#-------------------------------------- beta-analysis ----------------------------------------
## ---------- normalise Data --------------
```{r, echo = FALSE}
do_ter_cl_wide <- do_df_pl_tr_cl %>%
  group_by(library, genus_name) %>%
  summarise(reads = sum(sum_genus_count), .groups = "drop") %>%
  pivot_wider(names_from = genus_name, values_from = reads, values_fill = 0)

do_ter_cl_matrix <- do_ter_cl_wide %>%
  column_to_rownames("library") %>%
  as.matrix()

do_ter_cl_matrix_rel <- do_ter_cl_matrix / rowSums(do_ter_cl_matrix)
```
##--------- making PCoA ------------
```{r, echo = FALSE}
do_ter_cl_dist <- vegdist(do_ter_cl_matrix_rel, method = "bray")

pcoa_ter_cl <- pcoa(do_ter_cl_dist)
# calculate variance explained. 
do_var_explained <- 100 * pcoa_ter_cl$values$Relative_eig

# first only select asxis we want to plot (first two) and add right names to pcoa scores
pcoa_ter_cl_tibble <- as_tibble(pcoa_ter_cl$vectors[,1:2], .name_repair = "minimal") %>%
  mutate( library = rownames(do_ter_cl_matrix_rel))

# -------------------------------- make meta data ------------------------------
DOR2_libs <- do_df_pl_tr_cl %>%
  filter(core == "Core DOR 2") %>%
  distinct(library)
DOR3_libs <- do_df_pl_tr_cl %>%
  filter(core == "Core DOR 3") %>%
  distinct(library)
DOR4_libs <- do_df_pl_tr_cl %>%
  filter(core == "Core DOR 4") %>%
  distinct(library)

do_pcoa_meta <- pcoa_ter_cl_tibble %>%
  mutate(
    core = case_when(
      library %in% DOR2_libs$library ~ "Core DOR 2",
      library %in% DOR3_libs$library ~ "Core DOR 3",
      library %in% DOR4_libs$library ~ "Core DOR 4",
      TRUE ~ NA_character_
    )
  )

# --------------- add meta to pcoa -----------------------
pcoa_ter_cl_tibble <- pcoa_ter_cl_tibble %>%
  left_join(do_pcoa_meta %>% select(library, core), by = "library") %>%
  left_join(do_df_plant_ter %>% ungroup() %>% distinct(library, age_mean), by = "library")
            
  
```

##--- initial pcoa plot ---
```{r, echo = FALSE}

xlab1 <- sprintf("PCoA 1 (%.1f%%)", do_var_explained[1])
ylab1 <- sprintf("PCoA 2 (%.1f%%)", do_var_explained[2])

do_pcoa_pl1 <- ggplot(pcoa_ter_cl_tibble, aes(x = Axis.1, y = Axis.2)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 1) +
  geom_vline(xintercept = 0, color = "grey70", linewidth = 1) +
  geom_point(aes(fill = core), shape = 21, colour = "black", size = 3) +
  scale_x_continuous(limits = c(-0.5, 0.5)) +
  scale_y_continuous(limits = c(-0.5, 0.5)) +
  geom_label_repel(aes(label = age_mean), size = 3.5) +
  scale_fill_manual(values = do_site_colour, name = "Site", guide = "legend") +
  labs(x = xlab1, y = ylab1, title = "a") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_classic()

do_pcoa_pl1

ggsave("output/Do_pcoa_ter_cl1.pdf", plot = do_pcoa_pl1, width = 8, height = 6, dpi = 300)
```
## ------------- adding vectors -----------------------
```{r, echo = FALSE}
# get envfit vectors
do_fit_taxa <- envfit(pcoa_ter_cl$vectors, do_ter_cl_matrix_rel, permutations = 999 )

do_signif_taxa <- scores(do_fit_taxa, display = "vectors") %>%
  as.data.frame() %>%
  rownames_to_column("taxon") %>%
  filter(do_fit_taxa$vectors$pvals < 0.05) %>%
  mutate(
    r  = do_fit_taxa$vectors$r[do_fit_taxa$vectors$pvals < 0.05],
    r2 = r^2,
    p  = do_fit_taxa$vectors$pvals[do_fit_taxa$vectors$pvals < 0.05]
  ) %>%
  arrange(desc(r2)) %>%
  rename(genus_name = taxon)

```

## ------ add growth form data ------
```{r, echo = FALSE}
do_wcvp_native_genera <- do_df_raw %>%
  ungroup() %>%
  filter(genus_name %in% do_wcvp_native$genus)%>%
  distinct(genus_name) %>%
  left_join(do_wcvp_native %>% select(genus, lifeform_description), by = c("genus_name" = "genus"))

do_growth_form_native <- do_wcvp_native_genera %>%
  mutate(growth_form = case_when(
    # trees first
    str_detect(lifeform_description, "tree")                          ~ "tree",
    # shrubs
    str_detect(lifeform_description, "shrub")                         ~ "shrub",
    # aquatic/wetland
    str_detect(lifeform_description, "helophyte|hydro|aquatic")       ~ "aquatic/wetland",
    # climbers
    str_detect(lifeform_description, "climb|scrambl|liana")           ~ "climber",
    # grasses/sedges - based on genus since lifeform won't say "grass"
    genus_name %in% c("Carex", "Phragmites", "Poa", "Bromus", 
                      "Festuca", "Agrostis", "Holcus", "Alopecurus",
                      "Lolium", "Hordeum", "Triticum", "Avena", 
                      "Secale", "Oryza", "Zea", "Sorghum", "Panicum",
                      "Digitaria", "Echinochloa", "Setaria", "Cynodon",
                      "Eleusine", "Cyperus", "Schoenoplectus", 
                      "Eleocharis", "Typha", "Sparganium")             ~ "grass/sedge",
    # everything else is forb/herb
    str_detect(lifeform_description, "perennial|annual|biennial")     ~ "forb/herb",
    TRUE                                                               ~ "unknown"
  ))

# define priority order (1 = highest priority)
growth_form_priority <- c(
  "tree"             = 1,
  "shrub"            = 2,
  "grass/sedge"      = 3,
  "forb/herb"        = 4,
  "climber"          = 5,
  "aquatic/wetland"  = 6,
  "unknown"          = 7
)

# keep only highest priority growth form per genus
do_growth_form_native <- do_growth_form_native %>%
  mutate(priority = growth_form_priority[growth_form]) %>%
  group_by(genus_name) %>%
  slice_min(priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-priority)

do_wcvp_genera <- do_df_raw %>%
  ungroup() %>%
  filter(db == 'plant') %>%
  distinct(genus_name) %>%
  left_join(wcvp_names %>% select(genus, lifeform_description), by = c("genus_name" = "genus"))
  
do_growth_form_all <- do_wcvp_genera %>%
    mutate(growth_form = case_when(
    # handle NAs and empty strings first
    is.na(lifeform_description) | lifeform_description == "" ~ "unknown",
    
    # trees (before shrub so "shrub or tree" goes to tree... or swap if you prefer shrub)
    str_detect(lifeform_description, "tree")                          ~ "tree",
    
    # shrubs
    str_detect(lifeform_description, "shrub")                         ~ "shrub",
    
    # aquatic/wetland
    str_detect(lifeform_description, "helophyte|hydro|geophyte")      ~ "aquatic/wetland",
    
    # climbers/scramblers/lianas
    str_detect(lifeform_description, "climb|scrambl|liana|epiphyte")  ~ "climber",
    
    # grasses/sedges by genus
    genus_name %in% c("Carex", "Phragmites", "Poa", "Bromus",
                      "Festuca", "Agrostis", "Holcus", "Alopecurus",
                      "Lolium", "Hordeum", "Triticum", "Avena",
                      "Secale", "Oryza", "Zea", "Sorghum", "Panicum",
                      "Digitaria", "Echinochloa", "Setaria", "Cynodon",
                      "Eleusine", "Cyperus", "Schoenoplectus",
                      "Eleocharis", "Typha", "Sparganium",
                      "Stuckenia", "Potamogeton")                     ~ "grass/sedge",
    
    # forbs/herbs
    str_detect(lifeform_description, "perennial|annual|biennial|monocarpic") ~ "forb/herb",
    
    TRUE ~ "unknown"
  ))

do_growth_form_all <- do_growth_form_all %>%
  mutate(priority = growth_form_priority[growth_form]) %>%
  group_by(genus_name) %>%
  slice_min(priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-priority)

```

# ------------------------------ Check for FISH  -------------------------
```{r, echo = FALSE}
do_df_fish <- do_df_raw %>%
  filter(family_name %in% c("Cyprinidae", "Gobiidae", "Cichlidae", "Characidae", "Loricariidae", 
                            "Balitoridae", "Serranidae", "Labridae", "Scorpaenidae")) %>%
  slice_max(genus_count_assembly, n = 1, with_ties = FALSE) %>%
  select(library, core, depth, age_mean, phylum_name, order_name, family_name, genus_name, genus_count_assembly, sum_genus_count, deam5p_frac_genus, median_read_length_genus)

```


# -------------------- Library performance check ---------------------------
```{r, echo = FALSE}

do_lib_summary <- do_df_fm_gs %>%
  ungroup() %>%
  group_by(core, library) %>%
  summarise(
    depth_mean         = mean(depth, na.rm = TRUE),
    n_genera          = n_distinct(genus_name),
    n_genus_reads     = sum(sum_genus_count, na.rm = TRUE),
    n_family_reads    = sum(sum_family_count, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  group_by(core) %>%
  mutate(
    rank_genera      = rank(-n_genera),
    rank_genus_reads = rank(-n_genus_reads),
    rank_family_reads = rank(-n_family_reads),
    # combined rank score across all metrics
    combined_rank = rank_genera + rank_genus_reads + rank_family_reads
  ) %>%
  arrange(core, combined_rank)

do_lib_top3 <- do_lib_summary %>%
  group_by(core) %>%
  slice_min(combined_rank, n = 3)
  
do_lib_min3 <- do_lib_summary %>%
  group_by(core) %>%
  slice_max(combined_rank, n = 3)

do_lib_comb <- do_lib_summary %>%
  filter(library %in% c(do_lib_top3$library, do_lib_min3$library))
```
## show with polts
```{r, echo = FALSE}
do_performance_plot <- ggplot(do_lib_comb, aes(x = combined_rank, y = depth_mean, fill = core)) +
  geom_point(shape = 21, size = 4, colour = "black") +
  scale_y_reverse(limits = c(0, 180)) +
  scale_fill_manual(values = do_site_colour) +
  facet_wrap(~core, ncol = 3) +
  labs(x = "scored rank", y = "depth", title = "best and worse libs") +
  theme_classic() +
  theme(legend.position = "right")


do_performance_plot
ggsave("output/Do_library_performance_comb_rank.pdf", plot = do_performance_plot, width = 10, height = 6, dpi = 300)
```