---
title: "do_seq_deam_outliers"
author: "Tom"
date: "2026-09-21"
output: pdf_document
---
# Deamination outlier script
This script flags and stores assemblies which show suspicious (z score < -3) deamination rates. Its a across depth, weighted mean comparison. So assemlbies with more reads weigh heavier on the mean 'normal' deamination value at that depth. The interactive plotly plots show which assemblies are considered outliers (cross). Only the assemblies with a negative z-score crossing our threshold are considered contamination since higher deamination values indicate older material. 

The ridge plots serve as a way to visually check the deamination rate across depths. Long tails and skewed distributions are indications that something suss is going on.

# Loading
First load required libraries and data
```{r, echo=FALSE}
library(tidyverse)
library(patchwork)
library(ggplot2)
library(grid)
library(cowplot)
library(purrr)
library(fuzzyjoin)
library(openxlsx)
library(vegan)
library(ape)
library(ggrepel)
library(scico)
library(ggh4x)
library(slider)
library(stringr)
library(plotly)
library(ggridges)
```

# --------------------- simplify df and calculate w-mean -------------------
```{r, echo = FALSE}
# we only keep lake cores, only streptophyta, simplify core names and make identifier for long table, keep only relevant columns
do_la_deam <- do_df_fm_gs %>%
  ungroup() %>%
  filter(str_detect(core, "DOJ"),
         phylum_name == 'Streptophyta') %>%
  mutate(core = str_remove(core, "^Core "),
         identifier = paste(depth, core, genus_name, sep = "_")) %>%
  select("identifier", 2:4, "genus_name", "taxa_name", "genus_count_assembly", "sum_genus_count", "deam5p_frac_genus", "deam5p_genus", "ref_C_sum_5p_genus", "deam3p_frac_genus", "deam3p_genus", "ref_C_sum_3p_genus") %>%
  mutate(
    deam5p_frac_genus = if_else(ref_C_sum_5p_genus == 0, NA, deam5p_frac_genus),
    deam3p_frac_genus = if_else(ref_C_sum_3p_genus == 0, NA, deam3p_frac_genus)
  ) %>%
  group_by(core, genus_name)
  
# now we pivot longer to simplify coding and plotting since we have 3' and 5' deam which can now be combined. 
#weighted mean is calculated and z score of over 3 is flagged as an outlier 
do_la_deam_long <- do_la_deam %>%
  pivot_longer(c(deam3p_frac_genus, deam5p_frac_genus),
               names_to = "end", values_to = "frac") %>%
  group_by(core, depth, end) %>%
  mutate(
    w_mean = weighted.mean(frac, genus_count_assembly, na.rm = TRUE), 
    w_sd = sqrt(sum(genus_count_assembly * (frac - w_mean)^2, na.rm = TRUE) / sum(genus_count_assembly, na.rm = TRUE)),
    z_w = (frac - w_mean) / w_sd, 
    outlier = abs(z_w) > 3
  )

``` 
# -------------- plot deam plots with outlier indicators ---------------------
## ----------- for this we first make a fucntion to plot over all cores ----------
```{r, echo = FALSE}
# make function
make_deam_plot <- function(dat, core_name) {
  plot_ly() %>%
  add_markers(
    data = dat,
    x = ~depth,
    y = ~frac,
    color = ~end,
    symbols = c("circle", "x"),
    symbol = ~outlier,
    size = ~ifelse(outlier, 12, 8),
    text = ~paste(
      "identifier:", identifier,
      "<br>depth:", depth,
      "<br>", end, ":", round(frac, 3),
      "<br>Weighted Z:", round(z_w, 2),
      "<br>Outlier:", outlier
    ),
    hoverinfo = "text"
  ) %>%
  layout(
  title  = core_name,
  xaxis  = list(title = "Depth"),
  yaxis  = list(title = "Deamination fraction (genus)"),
  legend = list(title = list(text = "Series"))
)

}
 # apply function over list of plots
plots <- list()
for (a_core in unique(do_la_deam_long$core)) {
  plots[[a_core]] <- make_deam_plot(filter(do_la_deam_long, core == a_core), a_core)
}

plots
```





# ---------------------- Ridge Plots
now we make the ridge plots for a visual inspection of the deam distribution curves. shows if damage is consistent over depths, long tails towards low values indicate contamination, shows shift in deam over depth. Tells us if our z-score thing is workable since our approach works best if we have bell-curves. 

```{r, echo = FALSE}
make_ridge_plot <- function(dat, core_name) {
  ggplot(
    dat,
    aes(x = frac, y = factor(depth), fill = end, weight = genus_count_assembly)
  ) +
    geom_density_ridges(
      alpha = 0.7,
      scale = 3,
      rel_min_height = 0.01,
      color = "white"
    ) +
    labs(
      x = "Deamination fraction (genus)",
      y = "Depth",
      fill = "End",
      title = paste("Weighted distribution of deamination per depth:", core_name)
    ) +
    theme_minimal()
}

ridge_plots <- list()
for (a_core in unique(do_la_deam_long$core)) {
  ridge_plots[[a_core]] <- make_ridge_plot(
    filter(do_la_deam_long, core == a_core),
    a_core
  )
}

ridge_plots
```

# ---------------- save outliers ----------------
We only save outliers with negative z-score, since they have lower deamination than the 'normal'. save as .csv file to later call upon if we need to remove them for further analysis. We also check how many time that genus is flagged as outlier over the whole core. Also we remove the ones we now suspect as contamination (outlier AND frac < 0.1) we remove these from do_df_fm_gs and do_df_raw (creaitng the clean variants) to be used in our bleedover script

```{r, echo = FALSE}
# all outliers based on our criteria
do_la_outliers <- do_la_deam_long %>%
  filter(z_w < -2.99)
write.csv(do_la_outliers, "output/outliers_DOJ.csv", row.names = FALSE)



# also check how many times this genus is an outlier and how many times not. So maybe it is just variable in its deamination. f.e. if it is showing up as outlier 1/14 depths than likely that is just variance in deamination. If it shows up as outlier consistantly, and especially at lower depths, then this is likely contamination. 
# total depths this genus is present at in this core, and how many of those are outliers
genus_flag_rate <- do_la_deam_long %>%
  distinct(core, genus_name, depth, outlier) %>%     # one row per core-genus-depth-outlier status
  group_by(core, genus_name) %>%
  summarise(
    n_depths_total   = n_distinct(depth),
    n_depths_flagged = n_distinct(depth[outlier == TRUE]),
    frac_flagged     = n_depths_flagged / n_depths_total,
    .groups = "drop"
  )

# keep do_la_outliers exactly as-is, just add the fraction column
do_la_outliers_check <- do_la_outliers %>%
  left_join(genus_flag_rate, by = c("core", "genus_name"))

do_la_outliers_check


# for now we keep it simple and remove only that with less than 0.1 deamination of our flagged outliers. 
deam_outliers_to_remove <- do_la_outliers %>%
  filter(frac < 0.10) %>%
  distinct(core, depth, genus_name)

deam_outliers_to_remove

do_df_la_clean <- do_df_raw %>%
  filter(str_detect(core, "DOJ")) %>%
  anti_join(deam_outliers_to_remove, by = c("core", "depth", "genus_name"))

do_la_fm_gs_clean <- do_df_fm_gs %>%
  filter(str_detect(core, "DOJ"))  %>%
  anti_join(deam_outliers_to_remove, by = c("core", "depth", "genus_name"))

```

