---
title: "do_seq_bleedover_check"
author: "Tom"
date: "2026-09-23"
output: pdf_document
---

# Bleedover check
This script functions to pick up on assemlbies which show suspicious behaviour by maintaining consistent ratio's over depth. Meaning f.e. two genera show the same ratio throughout the whole core. This is suspect since real plants likelhy would vary over time. This similar ratio over a large depth indicates there is bleed-over from one to the other, so likely it is one single plant that produces the reads, not the two. To approach this a Pearson correlation test is run for every pair of genera. Genera which have a strong correlation (R^2 > 0.8) and are within the same order are plotted and saved to a .pdf for later inspection. The final verdict on bleed-over or not is done manually by inspecting the plots. 


# Load packages
```{r, echo = False}
library(readr)
library(ggpubr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(factoextra)
library(vegan)
library(viridis)
library(ggfortify)
library(plotly)
library(dplyr)
library(tidyr)
library(janitor)
library(readxl)
library(forcats)
library(ggplot2)
library(ggdendro)
library(vegan)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggforce)     
library(tidytext)
library(ggridges)

```

# Load data
Our data frame is already linked to age_depth model and simplified in Do_seq_pre_pro script.There we also already removed the duplicate genera, so but if you're working on species level you should rebuilid your df from the do_df_raw df. 
```{r, echo = FALSE}
# select phyla of interest _bo is bleed-over
# This df already is filtered for low read counts
do_df_bleed <- do_la_fm_gs_clean %>%
  ungroup() %>%
  filter(phylum_name == "Streptophyta",
         str_detect(core, "DOJ")) %>%
  mutate(core = str_remove(core, "^Core ")) %>% # get rid of Core 
  select(core, depth, age_mean, phylum_name, order_name, family_name, genus_name, taxa_name,
         genus_count_assembly)  # only keep relevant colums

# we also make a df without filtering for the low read counts
do_df_bleed_raw <- do_df_la_clean %>%
  ungroup() %>%
  group_by(core, depth, genus_name) %>%
  select(core, depth, age_mean, phylum_name, order_name, family_name, genus_name, taxa_name, genus_count_assembly) %>%
  filter(phylum_name == "Streptophyta",
         str_detect(core, "DOJ")) %>%
  mutate(core = str_remove(core, "^Core ")) %>% # get rid of Core 
  slice_max(order_by = genus_count_assembly, n = 1, with_ties= FALSE) %>%
  ungroup()
  
  

```
----------------------------- Inspect and select bleed-over taxa --------------

Here only use unfiltered readcount datset (so do_df_bleed_raw)

---------------------------------------------------------------------------------
## 0. Order by taxonomy
We make a factored list so plotting later is easy.
```{r, echo = FALSE}
genus_levels <- do_df_bleed_raw %>%
  distinct(phylum_name, order_name, family_name, genus_name) %>%
  arrange(phylum_name, order_name, family_name, genus_name) %>%
  pull(genus_name)

do_df_bleed_raw <- do_df_bleed_raw %>%
  mutate(genus_name = factor(genus_name, levels = genus_levels, ordered = TRUE))


# Now taxa selectino part. You only want taxa which will be in your final filtered df.

full_taxa_list <- levels(as.factor(do_df_bleed$genus_name))
write.csv(full_taxa_list, "output/full_taxa_list.csv")
# add an identifier so you see core and depth. and keep only genera that are in filtered df
do_df_bleed_raw <- do_df_bleed_raw %>%
  filter(genus_name %in% full_taxa_list) %>%
  ungroup() %>%
  group_by(core, depth) %>%
  mutate(identifier = paste(core, depth, sep = "_")) %>%
  ungroup()

```

--------------------------------------------------------------------------------------
###  0.1 Order check
If genera match to two orders this is bug, so we check this first. Flag if multiple order

```{r, echo = FALSE}
# genus order lookup
genus_order <- do_df_bleed_raw %>%
  select(genus_name, order_name) %>%
  distinct()

# now check if each genus maps to one order
bad_map <- genus_order %>%
  count(genus_name) %>%
  filter(n > 1)

if (nrow(bad_map) > 0) {
  stop("Some genera map to more than one order, check before proceeding")
}

```
--------------------------------------------------------------------------------------
## 1 & 2 Make wide community matrix AND Pairwise Pearson test
We'll do this as a function and then run the function over our list of cores within the main df. This is done in a nested loop. This is basically steps 1 (make wide community matrix) and step 2 (pairiwse pearson test) from Zoë's script but set under a loop so we do it for each core in our df. 
```{r, echo = FALSE}
bleedover_correlation <- function(dat) {
  # dat = data for one core
  comm_wide <- dat %>%
    group_by(identifier, genus_name) %>%
    summarise(genus_count_assembly = sum(genus_count_assembly), .groups = "drop") %>%
    pivot_wider(names_from = genus_name, values_from = genus_count_assembly, values_fill = 0)
  
  comm_mat <- comm_wide %>% select(-identifier) %>% as.matrix()
  genera <- colnames(comm_mat)
  
  if (ncol(comm_mat) < 2) return(NULL)  # if nog enough genera present to correlate
  
  results_list <- list() # this is a list which stores one comparison (genus 1 vs 2 f.e.) and stores r, r2 and p_value within one row. The whole for i and for j is just to check each combination once. 
  k <- 1 
  for (i in 1:(ncol(comm_mat) - 1)) {
    for (j in (i + 1):ncol(comm_mat)) {
      test <- cor.test(comm_mat[, i], comm_mat[, j], method = "pearson")
      r <- unname(test$estimate) 
      results_list[[k]] <- data.frame(
        genus_1 = genera[i], genus_2 = genera[j],
        r = r, r2 = r^2, p_value = test$p.value
      )
      k <- k + 1
    }
  }
  bind_rows(results_list) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH"))
}

# now loop it over each core in our df
bleedover_results <- list()
for (a_core in unique(do_df_bleed_raw$core)) {
  dat_core <- do_df_bleed_raw %>%
    filter(core == a_core) %>%
    mutate(identifier = paste(depth, core, sep = "_"))
  
  bleedover_results[[a_core]] <- bleedover_correlation(dat_core)
}

view(bleedover_results[[1]])

```

--------------------------------------------------------------------------------------
## 3 Add order info and filter > 0.8 R^2
Idea is to add order information and pull the pairs which have higher than 0.8 r2. From this we can make a list of suss taxa that are above 0.8 r2 AND have same order. 
We again do this in a loop since we got a list, not a separate df. 
```{r, echo = FALSE}

strong_same_order_list <- list()

for (a_core in names(bleedover_results)) {
  res <- bleedover_results[[a_core]]
  
  if(is.null(res)) next # empty entry gets skipped
  
  strong_same_order_list[[a_core]] <- res %>%
    left_join(genus_order, by = c("genus_1" = "genus_name")) %>%
    rename(order_1 = order_name) %>%
    left_join(genus_order, by = c("genus_2" = "genus_name")) %>%
    rename(order_2 = order_name) %>%
    filter(r2 > 0.8, order_1 == order_2) %>%
    arrange(desc(r2)) %>%
    mutate(core = a_core, .before = 1)   # <- add core column, put it first
}

view(strong_same_order_list[[3]])
```

---------------------------------------------------------------------------------------
## 4 Now plot the combo's per core in seperate pdfs. We use Zoë's code but loop it since we have list of results, not seperate per core. 
```{r, echo = FALSE}

plot_pair <- function(dat_wide, genus1, genus2, r2, p_value, p_adj, order_name = NA_character_) {
  dfp <- dat_wide %>%
    transmute(
      identifier = identifier,
      x = .data[[genus1]],
      y = .data[[genus2]]
    )
  
  ggplot(dfp, aes(x = x, y = y)) +
    geom_point(alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE) +
    theme_classic() +
    labs(
      x = genus1,
      y = genus2,
      title = paste0(genus1, " vs ", genus2,
                     if (!is.na(order_name)) paste0("  (", order_name, ")") else "")
    ) +
    annotate(
      "text",
      x = Inf, y = Inf,
      hjust = 1.05, vjust = 1.2,
      label = sprintf("R² = %.3f\np = %.2g\np_adj = %.2g", r2, p_value, p_adj)
    )
}

# loop over cores, one PDF per core
for (a_core in names(strong_same_order_list)) {
  
  pairs_to_plot <- strong_same_order_list[[a_core]]
  
  if (is.null(pairs_to_plot) || nrow(pairs_to_plot) == 0) next  # nothing flagged for this core
  
  # rebuild comm_wide for this core (needed for the raw x/y values to plot)
  dat_core <- do_df_bleed_raw %>%
    filter(core == a_core) %>%
    mutate(identifier = paste(depth, core, sep = "_"))
  
  comm_wide <- dat_core %>%
    group_by(identifier, genus_name) %>%
    summarise(genus_count_assembly = sum(genus_count_assembly), .groups = "drop") %>%
    pivot_wider(names_from = genus_name, values_from = genus_count_assembly, values_fill = 0)
  
  clean_name <- gsub("[ /]", "_", a_core)
  pdf(paste0("output/flagged_pairs_", clean_name, "_0.8.pdf"), width = 7, height = 6)
  
  for (i in seq_len(nrow(pairs_to_plot))) {
    p <- plot_pair(
      dat_wide  = comm_wide,
      genus1    = pairs_to_plot$genus_1[i],
      genus2    = pairs_to_plot$genus_2[i],
      r2        = pairs_to_plot$r2[i],
      p_value   = pairs_to_plot$p_value[i],
      p_adj     = pairs_to_plot$p_adj[i],
      order_name = pairs_to_plot$order_1[i]
    )
    print(p)
  }
  
  dev.off()
}
```
------------------------------------------------------------
------------------------------------------------------------------------------------------------------------------
# Remove exotics
We also want to flag genera which are not native to our area. These could either be bleedover, so the native taxa does not have a good reference genome resulting in our reads mapping to the exotic assembly which does have good coverage (f.e.). Or it could be contamination. For contamination we should expect the deamination to differ from our 'normal' deamination. 
First we find any exotic genera in our df using WCVP dtabase, this is already made in our Do_seq_pre_pro script. 

```{r, echo = FALSE}
do_native_taxa <- unique(do_wcvp_native$genus)
# problem, wcvp only does vascualr plants, so our mosses are also being flagged. For now just manually keep mosses in
mosses <- c("Sphagnum", "Polytrichum", "Grimmia", "Hypnum", "Anthoceros")

# match with genera in our df, single out exotics
do_exotic_candidates <- do_df_bleed_raw %>%
  distinct(genus_name) %>%
  filter(!genus_name %in% do_native_taxa) %>%
  filter(!genus_name %in% mosses)


write_csv(do_exotic_taxa, "output/do_exotic_taxa.csv")
```

## Add deamination scores
To see if this is likely a real cultivar brought in at the time or contamination we check their deamination scores based on the z-score made in the do_seq_deam_outlier script. 
```{r, echo = FALSE}
do_exotic_taxa <- do_la_deam_long %>%
  filter(genus_name %in% do_exotic_candidates$genus_name) %>%
  select(core, depth, identifier, genus_name, end, frac, z_w, outlier, genus_count_assembly) %>%
  arrange(core, depth, genus_name)

view(do_exotic_taxa)
```

------------------------------------------------------------------------------------------------------------------
# Merge exotics with bleedover correlations
An exotic genus that also correlates strongly with a native genus in the same order is good evidence it's a reference-mapping artifact rather than a real cultivar. We check, per core, which exotic candidates also show up in that core's flagged same-order pairs, and add those to exotic_bleedover_df. From this you can see what is exotic and from which taxa it is likely bleedover. 

```{r, echo = FALSE}
# for each core, find exotic candidates that appear in a flagged same-order pair,
# and record which genus they're correlated with
exotic_bleedover_list <- list()

for (a_core in names(strong_same_order_list)) {
  
  pairs <- strong_same_order_list[[a_core]]
  if (is.null(pairs) || nrow(pairs) == 0) next
  
  # long format: one row per genus per pair, so each genus's "partner" is easy to grab
  pairs_long <- bind_rows(
    pairs %>% transmute(genus = genus_1, related_genus = genus_2, r2, p_adj),
    pairs %>% transmute(genus = genus_2, related_genus = genus_1, r2, p_adj)
  )
  
  exotic_bleedover_list[[a_core]] <- pairs_long %>%
    filter(genus %in% do_exotic_candidates$genus_name) %>%
    transmute(
      core           = a_core,
      bleedover_genus = genus,
      related_genus,
      r2,
      p_adj,
      source = "exotic + bleedover correlated"
    )
}

exotic_bleedover_df <- bind_rows(exotic_bleedover_list)

exotic_bleedover_df
```


------------------------------------------------------------

# Determine actual bleedovers 
Now based on the plots we determine which ones we classify as actual bleedovers. BUT still have to decide if we throw these reads out or add them to the related genera? That's probably more fair. 
## Bleedovers

```{r, echo = FALSE}
# 2. Empty bleedover list, one slot per core, ready to fill in manually
bleedover_taxa <- setNames(
  vector("list", length(unique(do_df_bleed_raw$core))),
  unique(do_df_bleed_raw$core)
)
# here we fill in our bleedovers, they're empty for now
bleedover_taxa[["DOJ 1/24 center"]] <- c("Leonotis")
bleedover_taxa[["DOJ DEL 3/24/A"]]  <- c("Brainea", "Leonotis")
bleedover_taxa[["DOJ DEL 4/24/B"]]  <- c("Brainea", "Leonotis") 

```
# This is the final step where we make the lists from bleedover, aquatics, deamination and exotics to remove from our dataframe for analysis

## Remove true Aquatics
We don't want interference from what is growing in the lake. So we'll remove anything that grows submerged or floating. And we'll combine this with the bleedovers we identify from PDFs as one list so we can easily remove it before the analysis. 
```{r, echo = FALSE}
 #. Aquatics list - same across all cores
aquatics <- c("Myriophyllum", "Nuphar", "Potamogeton", "Stuckenia")
```

-----------------------------------------------------------------------------------------------------------------------------
# Taxa to Remove
Now final part is to remove our bleedovers, contamination based on deamination scores and our aquatics

```{r,echo = FALSE}

# build a per-core vector of genera to remove: bleedover_taxa genera + aquatics (every core)
taxa_to_remove <- map(bleedover_taxa, ~ unique(c(.x, aquatics)))


taxa_to_remove

# and remove them for the set we'll use for the classification step

do_df_class_list <- list()

for (a_core in unique(do_df_bleed$core)) {
  
  dat_core <- do_la_fm_gs_clean %>%
    filter(core == a_core)
  
  remove_list <- taxa_to_remove[[a_core]]   # NULL if this core has no entry in the list
  
  do_df_class_list[[a_core]] <- dat_core %>%
    filter(!genus_name %in% remove_list)
}

do_df_class <- bind_rows(do_df_class_list)
```

